---
title: "Advanced Usage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Advanced Usage

This guide covers advanced features and workflows for power users of ICellbioRpy.

## Multi-Language Support

ICellbioRpy supports both English and Chinese interfaces:

```{r language-config}
# Set language to Chinese
options(ICellbioRpy.lang = "zh")

# Set language to English (default)
options(ICellbioRpy.lang = "en")
```

## Smart Python Environment Management

### Interactive Environment Selection

The `smart_python_config()` function provides intelligent environment detection and selection:

```{r smart-python-config}
# Automatic detection and interactive selection
smart_python_config(verbose = TRUE, interactive = TRUE)

# Output example:
# 📋 Found multiple environments with anndata:
#   1. 1cellbio (anndata 0.12.0)
#   2. atlas (anndata 0.11.3)
#   3. scanpy (anndata 0.10.9)
# Please select environment to use (1-3):
```

If only one environment has anndata, it will be auto-selected. If multiple exist, you'll be prompted to choose.

### Multiple Python Environments

```{r multi-python-env}
# Configure specific environment for different projects
configure_python_env(conda_env = "scanpy_env", verbose = TRUE)

# Switch environments mid-session if needed
configure_python_env(conda_env = "another_env", verbose = TRUE)

# Check current environment status
check_anndata_available()
```

### Environment Validation

```{r env-validation}
# Comprehensive environment check
validate_environment <- function() {
  cat("=== Environment Validation ===\n")

  # Check Python configuration
  cat("Python environment: ")
  if (check_anndata_available()) {
    cat("✓ OK\n")
  } else {
    cat("✗ Failed\n")
    cat("Please run: configure_python_env(conda_env = 'your_env')\n")
    return(FALSE)
  }

  # Check required R packages
  required_packages <- c("Matrix", "Seurat", "SingleCellExperiment")
  for (pkg in required_packages) {
    cat("R package", pkg, ": ")
    if (requireNamespace(pkg, quietly = TRUE)) {
      cat("✓ OK\n")
    } else {
      cat("✗ Missing\n")
    }
  }

  return(TRUE)
}

# Run validation
validate_environment()
```

## Large Dataset Optimization

### Memory-Efficient Conversions

For large datasets, use these strategies to minimize memory usage:

```{r memory-efficient}
# Use direct ZIP to H5AD conversion (bypasses R objects)
iCellbio2H5ad(
  zip_file = "large_dataset.zip",
  h5ad_path = "output.h5ad",
  rownames = "gene_id",
  colnames = "cell_id"
)

# Specify data type detection for optimal performance
sce <- h5ad_to_sce("large_data.h5ad", use_x_as = "auto")
```

### Handling Multiple Assays

```{r multi-assays}
# Convert Seurat object with multiple assays
seurat_to_h5ad(
  seurat_obj,
  "output.h5ad",
  default_assay = "RNA",           # Main assay for X matrix
  layer = "data",                  # Use normalized data
  include_reductions = TRUE        # Include PCA, UMAP, etc.
)

# All other assays will be stored as layers in the H5AD file
```

## Stereo-seq Spatial Transcriptomics

### Reading Large GEF Files

For large Stereo-seq GEF files, use memory-optimized reading strategies:

```{r large-gef}
# Memory-optimized reading with cell limit
stereo_subset <- read_gef(
  "large_sample.cellbin.gef",
  max_cells = 5000,                           # Limit cell count
  region = c(1000, 5000, 1000, 5000),        # Spatial region filter
  gene_list = c("EPCAM", "KRT8", "VIM")      # Gene filter
)

# Read specific spatial region only
stereo_region <- read_gef(
  "sample.gef",
  region = c(1000, 3000, 1000, 3000),  # minX, maxX, minY, maxY
  include_cellborder = TRUE
)
```

### Direct GEF to H5AD Conversion

```{r gef-to-h5ad-advanced}
# Direct conversion with full options (memory efficient)
gef_to_h5ad(
  gef_file = "sample.cellbin.gef",
  h5ad_file = "output.h5ad",
  bin_type = "cell_bins",
  include_cellborder = TRUE,    # cell_borders stored in adata.uns
  include_spatial = TRUE,       # spatial coordinates stored in adata.obsm
  max_cells = 10000,            # memory management
  region = c(1000, 5000, 1000, 5000),      # spatial region filter
  gene_list = c("GAPDH", "ACTB", "CD45"),  # gene filter
  overwrite = TRUE
)
```

### Spatial Visualization with Customization

```{r spatial-viz-advanced}
# Advanced spatial visualization with cell borders
plot_cells_with_borders(
  stereo_data,
  color_by = "cluster",        # Color by cluster or gene expression
  show_borders = TRUE,         # Show cell boundaries
  border_color = "white",      # Border color
  border_size = 0.5,           # Border line width
  point_size = 1,              # Cell point size
  alpha = 0.8                  # Transparency
)

# Color by gene expression
plot_cells_with_borders(
  stereo_data,
  color_by = "EPCAM",
  show_borders = TRUE,
  border_color = "black"
)
```

### Using Square Bins (Without Cell Segmentation)

```{r stereo-bins}
# Use square bins instead of cell segmentation
stereo_bins <- read_gef(
  "sample.gef",
  bin_type = "bins",
  bin_size = 50  # 50x50 pixel bins
)

# Convert bins to H5AD
gef_to_h5ad(
  "sample.gef",
  "bins_output.h5ad",
  bin_type = "bins",
  bin_size = 50
)
```

## GMT Gene Set Batch Processing

### Performance Optimization with Pre-built Tables

For processing multiple GMT files, pre-build gene lookup tables for significant performance gains:

```{r gmt-batch-advanced}
# Step 1: Pre-build lookup tables for all species
prebuild_gene_lookup_tables(
  data_dir = "~/gene_mapping_data",
  output_file = "gene_lookup_tables.rdata"
)

# Step 2: Load into global environment (creates master_lookup_tables variable)
load("gene_lookup_tables.rdata")

# Step 3: Batch process multiple GMT files
gmt_files <- list.files(pattern = "\\.gmt$")
for (gmt in gmt_files) {
  output_dir <- gsub("\\.gmt$", "_gesel", gmt)
  preprocess_gmt_custom(
    gmt_file = gmt,
    species = "9606",          # or "10090" for mouse
    output_dir = output_dir
  )
  cat("Processed:", gmt, "\n")
}
```

### Custom Gene Mapping Files

```{r gmt-custom}
# Use with custom gene mapping files
preprocess_gmt_with_custom_mapping(
  gmt_file = "custom_pathways.gmt",
  species = "10090",  # Mouse
  gene_mapping_files = list(
    symbol = "mouse_symbols.tsv",
    entrez = "mouse_entrez.tsv",
    ensembl = "mouse_ensembl.tsv"
  ),
  collection_name = "mouse_pathways",
  collection_desc = "Custom mouse pathways",
  output_dir = "mouse_gesel",
  auto_download_missing = TRUE  # Auto-download missing mapping files
)
```

### Supported Species

| Species ID | Species Name |
|------------|--------------|
| 9606 | Human (Homo sapiens) |
| 10090 | Mouse (Mus musculus) |
| 10116 | Rat (Rattus norvegicus) |
| 7227 | Fruit fly (Drosophila melanogaster) |
| 6239 | Nematode (Caenorhabditis elegans) |
| 7955 | Zebrafish (Danio rerio) |
| 9598 | Chimpanzee (Pan troglodytes) |

## Batch Processing Workflows

### Processing Multiple 1CellBio Files

```{r batch-1cellbio}
# Function to process multiple ZIP files
process_multiple_1cellbio <- function(zip_files, output_dir) {
  for (zip_file in zip_files) {
    # Extract sample name from filename
    sample_name <- tools::file_path_sans_ext(basename(zip_file))
    output_h5ad <- file.path(output_dir, paste0(sample_name, ".h5ad"))

    # Convert to H5AD with auto-detection
    iCellbio2H5ad(
      zip_file = zip_file,
      h5ad_path = output_h5ad,
      overwrite = FALSE,
      name_conflict = "make_unique"
    )
  }
}

# Usage
zip_files <- list.files("input_dir", pattern = "\\.zip$", full.names = TRUE)
process_multiple_1cellbio(zip_files, "output_h5ad")
```

### Advanced 10X Integration

```{r advanced-10x}
# Create detailed sample metadata
sample_info <- data.frame(
  Sample_id = c("ctrl_1", "ctrl_2", "treat_1", "treat_2"),
  condition = c("control", "control", "treatment", "treatment"),
  batch = c("batch1", "batch2", "batch1", "batch2"),
  mtx_fns = c("ctrl1/matrix.mtx.gz", "ctrl2/matrix.mtx.gz",
              "treat1/matrix.mtx.gz", "treat2/matrix.mtx.gz"),
  features_fns = c("ctrl1/features.tsv.gz", "ctrl2/features.tsv.gz",
                   "treat1/features.tsv.gz", "treat2/features.tsv.gz"),
  barcodes_fns = c("ctrl1/barcodes.tsv.gz", "ctrl2/barcodes.tsv.gz",
                   "treat1/barcodes.tsv.gz", "treat2/barcodes.tsv.gz")
)

# Integrate with custom QC thresholds
read_10x_mtx_to_h5ad(
  csv_file = "detailed_samples.csv",
  output_h5ad = "integrated_with_metadata.h5ad",
  min_counts_per_cell = 500,  # Stricter QC
  verbose = TRUE
)
```

### Batch Stereo-seq Processing

```{r batch-stereo}
# Process multiple GEF files
gef_files <- list.files(pattern = "\\.gef$", full.names = TRUE)

for (gef_file in gef_files) {
  output_name <- gsub("\\.gef$", ".h5ad", basename(gef_file))

  # Direct conversion with memory management
  gef_to_h5ad(
    gef_file = gef_file,
    h5ad_file = file.path("output", output_name),
    bin_type = "cell_bins",
    include_cellborder = TRUE,
    include_spatial = TRUE,
    max_cells = 10000,  # Memory limit
    overwrite = FALSE
  )

  cat("Converted:", gef_file, "\n")
}
```

## Error Handling and Troubleshooting

### Robust File Processing

```{r error-handling}
# Safe conversion with error handling
safe_convert_to_h5ad <- function(zip_file, output_file) {
  tryCatch({
    iCellbio2H5ad(
      zip_file = zip_file,
      h5ad_path = output_file,
      overwrite = TRUE,
      name_conflict = "make_unique"
    )
    message("✓ Successfully converted: ", basename(zip_file))
  }, error = function(e) {
    warning("✗ Failed to convert ", basename(zip_file), ": ", e$message)
    return(NULL)
  })
}
```

### Column Auto-Detection Fallback

```{r column-detection-advanced}
# When column detection fails, auto-selects first column with bilingual warning
seurat_obj <- as.Seurat.1CB(data)  # auto-detects columns

# View available columns if auto-detection is unclear
show_column_options(data)

# Manually specify if needed
seurat_obj <- as.Seurat.1CB(data,
                           rownames = "gene_id",
                           colnames = "cell_id",
                           auto_detect = FALSE)
```

### Name Conflict Strategies

```{r name-conflicts-advanced}
# Strict mode - error on duplicates
tryCatch({
  seurat_obj <- as.Seurat.1CB(
    data,
    rownames = "gene_id",
    colnames = "cell_id",
    name_conflict = "error"
  )
}, error = function(e) {
  message("Duplicate names detected, switching to make_unique mode")
  seurat_obj <- as.Seurat.1CB(
    data,
    rownames = "gene_id",
    colnames = "cell_id",
    name_conflict = "make_unique"
  )
})
```

## Performance Optimization

### Sparse Matrix Best Practices

```{r sparse-optimization}
# Monitor memory usage during conversion
monitor_conversion <- function(input_file, output_file) {
  start_time <- Sys.time()
  start_memory <- gc()

  # Perform conversion
  result <- h5ad_to_seurat(input_file, verbose = TRUE)

  end_time <- Sys.time()
  end_memory <- gc()

  cat("Conversion completed in:",
      round(difftime(end_time, start_time, units = "secs"), 2), "seconds\n")

  return(result)
}
```

### Memory Management for Large Files

```{r memory-management}
# Set memory limits for large operations
options(future.globals.maxSize = 8000 * 1024^2)  # 8GB

# Process in chunks if dataset is very large
process_large_dataset <- function(large_file, chunk_size = 10000) {
  # Read subset of cells
  data <- read_gef(large_file, max_cells = chunk_size)

  # Process subset
  # ... your analysis ...

  # Move to next chunk
}
```

## Integration with Other Packages

### Scanpy Integration

```{r scanpy-integration}
# After creating H5AD file, use with scanpy
# This would be run in Python:
#
# import scanpy as sc
# adata = sc.read_h5ad("output.h5ad")
#
# # Access spatial coordinates
# spatial_coords = adata.obsm['spatial']
#
# # Access cell borders
# cell_borders = adata.uns['cell_borders']
#
# sc.pp.highly_variable_genes(adata)
# sc.tl.pca(adata)
# sc.pl.pca_variance_ratio(adata)
```

### Bioconductor Integration

```{r bioconductor}
# Convert to SingleCellExperiment for Bioconductor workflows
sce <- h5ad_to_sce("data.h5ad")

# Use with scater, scran, etc.
# library(scater)
# sce <- addPerCellQC(sce)
# plotColData(sce, "total")
```

### Seurat Spatial Workflows

```{r seurat-spatial}
# Convert Stereo-seq data to Seurat
stereo_data <- read_gef("sample.cellbin.gef", include_cellborder = TRUE)
seurat_obj <- as.Seurat(stereo_data)

# Use with Seurat's spatial features
# Spatial coordinates in @reductions$spatial
# Cell borders in @misc$cell_borders

# Standard Seurat workflow continues
seurat_obj <- NormalizeData(seurat_obj)
seurat_obj <- FindVariableFeatures(seurat_obj)
seurat_obj <- ScaleData(seurat_obj)
```

## Custom Workflows

### Building Analysis Pipelines

```{r custom-pipeline}
# Complete analysis pipeline for 1CellBio data
run_1cellbio_pipeline <- function(zip_file, output_dir, analysis_type = "seurat") {
  # Step 1: Convert to H5AD
  h5ad_file <- file.path(output_dir, "data.h5ad")
  iCellbio2H5ad(zip_file, h5ad_file,
                overwrite = FALSE,
                name_conflict = "make_unique")

  # Step 2: Convert to analysis format
  if (analysis_type == "seurat") {
    obj <- h5ad_to_seurat(h5ad_file)
    saveRDS(obj, file.path(output_dir, "seurat_object.rds"))
  } else if (analysis_type == "sce") {
    obj <- h5ad_to_sce(h5ad_file)
    saveRDS(obj, file.path(output_dir, "sce_object.rds"))
  }

  return(obj)
}

# Usage
run_1cellbio_pipeline("sample.zip", "output/", "seurat")
```

### Spatial Transcriptomics Pipeline

```{r spatial-pipeline}
# Complete Stereo-seq analysis pipeline
run_stereo_pipeline <- function(gef_file, output_dir) {
  # Step 1: Read GEF file with optimizations
  stereo_data <- read_gef(
    gef_file,
    bin_type = "cell_bins",
    include_cellborder = TRUE,
    max_cells = 50000
  )

  # Step 2: Convert to H5AD for Python analysis
  h5ad_file <- file.path(output_dir, "spatial_data.h5ad")
  stereo_to_h5ad(stereo_data, h5ad_file, include_spatial = TRUE)

  # Step 3: Convert to Seurat for R analysis
  seurat_obj <- as.Seurat(stereo_data)

  # Step 4: Create spatial visualization
  plot_cells_with_borders(
    stereo_data,
    color_by = "cluster",
    show_borders = TRUE,
    border_color = "white"
  )

  return(list(stereo_data = stereo_data,
              seurat_obj = seurat_obj,
              h5ad_file = h5ad_file))
}

# Usage
results <- run_stereo_pipeline("sample.cellbin.gef", "output/")
```

## Troubleshooting Common Issues

### Issue 1: Python Environment Problems

```{r troubleshoot-python}
# If you get "anndata not found" errors:
# 1. Use smart interactive configuration
smart_python_config(verbose = TRUE, interactive = TRUE)

# 2. Or manually check conda environments
system("conda env list")

# 3. Activate the correct environment and install anndata:
# conda activate your_env
# pip install anndata

# 4. Reconfigure in R:
configure_python_env(conda_env = "your_env", verbose = TRUE)
```

### Issue 2: Memory Issues with Large Files

```{r troubleshoot-memory}
# For very large files, use direct conversion:
# Avoid: data -> Seurat -> H5AD
# Use: ZIP -> H5AD directly
iCellbio2H5ad(zip_file, h5ad_file,
             overwrite = FALSE,
             name_conflict = "make_unique")

# For GEF files, use region filtering
gef_to_h5ad("large.gef", "output.h5ad",
            region = c(1000, 5000, 1000, 5000),
            max_cells = 10000)
```

### Issue 3: Column Detection Failures

```{r troubleshoot-columns}
# View available columns
show_column_options(data)

# Use auto-selection with warning (recommended)
seurat_obj <- as.Seurat.1CB(data)  # Auto-selects first column

# Or manually specify
seurat_obj <- as.Seurat.1CB(data,
                           rownames = "gene_id",
                           colnames = "cell_id",
                           auto_detect = FALSE)
```

### Issue 4: Duplicate Names

```{r troubleshoot-names}
# Always use make_unique for robustness:
obj <- as.Seurat.1CB(data,
                     rownames = "gene_id",
                     colnames = "cell_id",
                     name_conflict = "make_unique")
```

### Issue 5: GMT Processing Slow

```{r troubleshoot-gmt}
# Pre-build lookup tables for batch processing
prebuild_gene_lookup_tables(
  data_dir = "~/gene_mapping_data",
  output_file = "gene_lookup_tables.rdata"
)

load("gene_lookup_tables.rdata")

# Now process GMT files - much faster
preprocess_gmt_custom("pathways.gmt", species = "9606")
```
